Sufficient Conditions for Reachability in Automata Networks with Priorities

### 2014-12-18 parts/example.tex
refactor successive reachability in metazoan example

\subsection{Flattening of a model: Metazoan segmentation}

In this section, we illustrate the flattening method presented in \pref{ssec:flattening}
on a model of metazoan segmentation inspired from a first translation to Process Hitting given in \cite{PMR10-TCSB}.
This model was originally established in silico in \cite{MSB:MSB4100192} in a differential equations framework.
It is composed of a wavefront gene $f$ that activates the gap-gene $a$ whose products are responsible for stripes formation.
Gene $f$ also activates a gene $c$ whose products represses the gene $a$.
The auto-inhibition of $c$ generalises a chain of repressors on $a$.
The auto-inhibition of $f$, which terminates the stripes formation in the original model,
has been removed in order to focus on the stationary dynamics of the model.

@@

\begin{table}[p]
\modMF{
\begin{align*}
  \PHh^{(1)} = \{ \quad
    & \PHfrappe{f_1}{fc_{00}}{fc_{10}} \quad , \quad
    \PHfrappe{f_1}{fc_{01}}{fc_{11}} \quad , \\
    & \PHfrappe{f_0}{fc_{10}}{fc_{00}} \quad , \quad
    \PHfrappe{f_0}{fc_{11}}{fc_{01}} \quad , \\
    & \PHfrappe{c_1}{fc_{00}}{fc_{01}} \quad , \quad
    \PHfrappe{c_1}{fc_{10}}{fc_{11}} \quad , \\
    & \PHfrappe{c_0}{fc_{01}}{fc_{00}} \quad , \quad
    \PHfrappe{c_0}{fc_{11}}{fc_{10}}
  \quad \} \\
  \PHh^{(2)} = \{ \quad
    & \PHfrappe{c_1}{a_1}{a_0} \quad , \quad
    \PHfrappe{fc_{10}}{a_0}{a_1}
  \quad \} \\
  \PHh^{(3)} = \{ \quad
    & \PHfrappe{c_1}{c_1}{c_0} \quad , \quad
    \PHfrappe{f_1}{c_0}{c_1}
  \quad \} \\
\end{align*}%
}%
  \caption{\label{tab:metazoan-php-actions}%
    Sets of actions of the metazoan segmentation model depicted
    in \pref{fig:metazoan-php}.
  }
\end{table}

\pref{fig:metazoan-ph} gives the flattening of this model,
\modMF{
whose actions are detailed in \pref{tab:metazoan-ph-actions}.
}%
It contains $2$ classes of priority, and prioritised actions are only used to update cooperative sorts.
Note that according to the simplifications mentioned at the end of \pref{ssec:flattening},
the auto-action $h_1 = \PHhit{c_1}{c_1}{c_0}$ in $\oPH$ has been flattened to a regular action $\PHhit{a_0}{c_1}{c_0}$
because its playability property could be simplified to $\Fop{h_1} \equiv c_1 \wedge a_0 \equiv a_0$.

The static analysis results of \pref{ssec:ua} and \pref{ssec:ordered-ua} can be used to check if the model is functional, \ie
if genes $a$ and $c$ are successively activated and deactivated,
thus leading to the formation of regular stripes.
Starting from context $\ctx = \PHstate{f_1, a_0, c_0}$,
we want to check the reachability of
$\w = \PHobj{a_0}{a_1} \PHconcat \PHobj{c_0}{c_1} \PHconcat \PHobj{a_1}{a_0} \PHconcat \PHobj{c_1}{c_0}$.
Rather than checking this whole objective sequence,
we decompose it into the following ones:
\begin{align*}
  \ctx^1 &= \PHstate{f_1, a_0, c_0} \text{ and } P_1 = \PHobj{a_0}{a_1} &
  \ctx^3 &= \PHstate{f_1, a_1, c_1} \text{ and } P_2 = \PHobj{a_1}{a_0} \\
  \ctx^2 &= \PHstate{f_1, a_1, c_0} \text{ and } P_3 = \PHobj{c_0}{c_1} &
  \ctx^4 &= \PHstate{f_1, a_0, c_1} \text{ and } P_4 = \PHobj{c_1}{c_0}
\end{align*}
Given \pref{th:approxinf}, we have: $\forall i \in \segm{1}{4}, \muconcr_{\ctx^i}(P_i) \neq \emptyset$.
Finally, after checking that it can be applied for $i = 3$ down to $i = 1$,
\pref{thm:ordered-ua} gives: $\uconcr(\w) \neq \emptyset$.
Thus the model is functional.

\todo{Le résultat peut être raffiné avec l'atteignabilité successive}

@@

\begin{table}[p]
\modMF{
\begin{align*}
  \PHh^{(1)} = \{ \quad
    & \PHfrappe{f_1}{fc_{00}}{fc_{10}} \quad , \quad
    \PHfrappe{f_1}{fc_{01}}{fc_{11}} \quad , \\
    & \PHfrappe{f_0}{fc_{10}}{fc_{00}} \quad , \quad
    \PHfrappe{f_0}{fc_{11}}{fc_{01}} \quad , \\
    & \PHfrappe{c_1}{fc_{00}}{fc_{01}} \quad , \quad
    \PHfrappe{c_1}{fc_{10}}{fc_{11}} \quad , \\
    & \PHfrappe{c_0}{fc_{01}}{fc_{00}} \quad , \quad
    \PHfrappe{c_0}{fc_{11}}{fc_{10}} \quad , \\
    & \PHfrappe{f_1}{fa_{00}}{fa_{10}} \quad , \quad
    \PHfrappe{f_1}{fa_{01}}{fa_{11}} \quad , \\
    & \PHfrappe{f_0}{fa_{10}}{fa_{00}} \quad , \quad
    \PHfrappe{f_0}{fa_{11}}{fa_{01}} \quad , \\
    & \PHfrappe{a_1}{fa_{00}}{fa_{01}} \quad , \quad
    \PHfrappe{a_1}{fa_{10}}{fa_{11}} \quad , \\
    & \PHfrappe{a_0}{fa_{01}}{fa_{00}} \quad , \quad
    \PHfrappe{a_0}{fa_{11}}{fa_{10}}
  \quad \} \\
  \PHh^{(2)} = \{ \quad
    & \PHfrappe{c_1}{a_1}{a_0} \quad , \quad
    \PHfrappe{fc_{10}}{a_0}{a_1} \quad , \\
    & \PHfrappe{a_0}{c_1}{c_0} \quad , \quad
    \PHfrappe{fa_{11}}{c_0}{c_1}
  \quad \}
\end{align*}%
}%
  \caption{\label{tab:metazoan-ph-actions}%
    Sets of actions of the flattened metazoan segmentation model depicted
    in \pref{fig:metazoan-ph}.
  }
\end{table}

\subsection{Large-scale Application: T-cell receptor}

In order to support the scalability and applicability of our under-approximation of reachability, we
apply our new approach for the analysis of large-scale model of the T-cell receptor (TCR)
signalling pathway \cite{tcrsig94}.
This model gathers 94 interacting components and is specified as a Boolean network.
The under-approximation presented in this paper has been implemented in the existing Pint
software\footnote{Pint is freely available at \url{http://loicpauleve.name/pint}.}.

>>>>

\modMF{
This section aims at giving application examples of the static analysis method
that we developed in \pref{sec:sa}.
In \pref{ssec:ex-metazoan}, we apply our method to a small model of the
metazoan segmentation process,
and demonstrate how the flattening of \pref{ssec:flattening}
be used in such a case.
In \label{ssec:ex-tcrsig}, we apply our method to a large-scale model
of the T-cell receptor signalling pathway
in order to show the scalability of our method.
}

\subsection{Flattening of a model: Metazoan segmentation}
\label{ssec:ex-metazoan}

\modMF{
We give in this section a detailed example of the application of the sequential
under-approximation proposed in \pref{ssec:ordered-ua},
which consists of several applications of the methods developed in
\pref{ssec:ua}.
For this, we first have to use the flattening method presented in \pref{ssec:flattening}
because the considered model contains 3 classes of priorities (and not only 2, as required).
}

We consider in the following a model of metazoan segmentation
inspired from a first translation to Process Hitting given in \cite{PMR10-TCSB}.
This model was originally established in silico in \cite{MSB:MSB4100192} in a differential equations framework.
It is composed of a wavefront gene $f$ that activates the gap-gene $a$ whose products are responsible for stripes formation.
Gene $f$ also activates a gene $c$ whose products represses the gene $a$.
The auto-inhibition of $c$ generalises a chain of repressors on $a$.
The auto-inhibition of $f$, which terminates the stripes formation in the original model,
has been removed in order to focus on the stationary dynamics of the model.

@@

\begin{table}[p]
\modMF{
\begin{align*}
  \PHh^{(1)} = \{ \quad
    & \PHfrappe{f_1}{fc_{00}}{fc_{10}} \quad , \quad
    \PHfrappe{f_1}{fc_{01}}{fc_{11}} \quad , \\
    & \PHfrappe{f_0}{fc_{10}}{fc_{00}} \quad , \quad
    \PHfrappe{f_0}{fc_{11}}{fc_{01}} \quad , \\
    & \PHfrappe{c_1}{fc_{00}}{fc_{01}} \quad , \quad
    \PHfrappe{c_1}{fc_{10}}{fc_{11}} \quad , \\
    & \PHfrappe{c_0}{fc_{01}}{fc_{00}} \quad , \quad
    \PHfrappe{c_0}{fc_{11}}{fc_{10}}
  \quad \} \\
  \PHh^{(2)} = \{ \quad
    & \PHfrappe{c_1}{a_1}{a_0} \quad , \quad
    \PHfrappe{fc_{10}}{a_0}{a_1}
  \quad \} \\
  \PHh^{(3)} = \{ \quad
    & \PHfrappe{c_1}{c_1}{c_0} \quad , \quad
    \PHfrappe{f_1}{c_0}{c_1}
  \quad \} \\
\end{align*}%
}%
  \caption{\label{tab:metazoan-php-actions}%
    The sets of actions of the metazoan segmentation model depicted
    in \pref{fig:metazoan-php}.
  }
\end{table}

\pref{fig:metazoan-ph} gives the flattening of this model,
\modMF{
whose actions are detailed in \pref{tab:metazoan-ph-actions}.
}%
It contains $2$ classes of priority, and prioritised actions are only used to update cooperative sorts.
Note that according to the simplifications mentioned at the end of \pref{ssec:flattening},
the auto-action $h_1 = \PHhit{c_1}{c_1}{c_0}$ in the original model has been flattened to a regular action $\PHhit{a_0}{c_1}{c_0}$
because its playability property could be simplified to $\Fop{h_1} \equiv c_1 \wedge a_0 \equiv a_0$.

@@

\begin{table}[p]
\modMF{
\begin{align*}
  \PHh^{(1)} = \{ \quad
    & \PHfrappe{f_1}{fc_{00}}{fc_{10}} \quad , \quad
    \PHfrappe{f_1}{fc_{01}}{fc_{11}} \quad , \\
    & \PHfrappe{f_0}{fc_{10}}{fc_{00}} \quad , \quad
    \PHfrappe{f_0}{fc_{11}}{fc_{01}} \quad , \\
    & \PHfrappe{c_1}{fc_{00}}{fc_{01}} \quad , \quad
    \PHfrappe{c_1}{fc_{10}}{fc_{11}} \quad , \\
    & \PHfrappe{c_0}{fc_{01}}{fc_{00}} \quad , \quad
    \PHfrappe{c_0}{fc_{11}}{fc_{10}} \quad , \\
    & \PHfrappe{f_1}{fa_{00}}{fa_{10}} \quad , \quad
    \PHfrappe{f_1}{fa_{01}}{fa_{11}} \quad , \\
    & \PHfrappe{f_0}{fa_{10}}{fa_{00}} \quad , \quad
    \PHfrappe{f_0}{fa_{11}}{fa_{01}} \quad , \\
    & \PHfrappe{a_1}{fa_{00}}{fa_{01}} \quad , \quad
    \PHfrappe{a_1}{fa_{10}}{fa_{11}} \quad , \\
    & \PHfrappe{a_0}{fa_{01}}{fa_{00}} \quad , \quad
    \PHfrappe{a_0}{fa_{11}}{fa_{10}}
  \quad \} \\
  \PHh^{(2)} = \{ \quad
    & \PHfrappe{c_1}{a_1}{a_0} \quad , \quad
    \PHfrappe{fc_{10}}{a_0}{a_1} \quad , \\
    & \PHfrappe{a_0}{c_1}{c_0} \quad , \quad
    \PHfrappe{fa_{11}}{c_0}{c_1}
  \quad \}
\end{align*}%
}%
  \caption{\label{tab:metazoan-ph-actions}%
    The sets of actions of the flattened metazoan segmentation model depicted
    in \pref{fig:metazoan-ph}.
  }
\end{table}

\modMF{
Finally, the static analysis results presented in
\pref{ssec:ua} and \pref{ssec:ordered-ua} can be used to check if the model is functional,
\ie if gene $a$ can oscillate, thus leading to the formation of stripes.
Starting from context $\ctx = \PHstate{f_1, a_0, c_0, fc_{10}, fa_{00}}$,
we thus want to check the reachability of
$\w = \PHobj{a_0}{a_1} \PHconcat \PHobj{a_1}{a_0} \PHconcat \PHobj{a_0}{a_1}$.
Rather than checking this whole objective sequence with \pref{th:approxinf},
we break it using \pref{thm:ordered-ua}.
We first build the first saturated graph of local causality
based on the first objective, $\w_1 = \PHobj{a_0}{a_1}$, in the initial context $\ctx$.
This graph is depicted in \pref{fig:metazoan-sa}(left).
%and given \pref{tmh:approxinf}, we have:
We have especially:
\[\allprocs(\thisB{\w_1}{\ctx}) = \{ a_0, c_0, f_1, fc_{10} \}\]
\[\lastprocs(\PHobj{a_0}{a_1}) = \{\{ a_1, c_0, f_1, fc_{10} \}\}\]
Therefore, the second step consists of computing the saturated graph of local causality
of the objective $\w_2 = \PHobj{a_1}{a_0}$
in the context
\[\ctx' = \ctx \Cap \allprocs(\thisB{\w_1}{\ctx}) \Cap \lastprocs(\PHobj{a_0}{a_1}) = \PHstate{a_1, c_0, f_1, fc_{10}, fa_{00}}\]
This second graph is given in \pref{fig:metazoan-sa}(middle).
Finally, because
\[\allprocs(\thisB{\w_2}{\ctx'}) = \{ a_1, c_1, f_1, fa_{11} \}\]
\[\lastprocs(\PHobj{a_1}{a_0}) = \{\{ a_0, c_1 \}\}\]
The last step is then checking 
the objective $\w_3 = \PHobj{a_0}{a_1}$
in the context
\[\ctx'' = \ctx' \Cap \allprocs(\thisB{\w_2}{\ctx'}) \Cap \lastprocs(\PHobj{a_1}{a_0}) = \PHstate{a_0, c_1, f_1, fc_{10}, fa_{11}}\]
This last graph is depicted in 
As all the produced \pref{fig:metazoan-sa}(right).
In conclusion, by using \pref{thm:ordered-sa} on these saturated graphs of local causality,
we can conclude that the initial objective $\w$ is reachable,
and thus that the PH model of \pref{fig:metazoan-ph} is functional.
This result can be extended to the model of \pref{fig:metazoan-php}
because they have the same dynamics (given \pref{th:bisimPHP}).
}

% $\w = \PHobj{a_0}{a_1} \PHconcat \PHobj{c_0}{c_1} \PHconcat \PHobj{a_1}{a_0} \PHconcat \PHobj{c_1}{c_0}$.
% Rather than checking this whole objective sequence,
% we decompose it into the following ones:
% \begin{align*}
%   \ctx^1 &= \PHstate{f_1, a_0, c_0} \text{ and } P_1 = \PHobj{a_0}{a_1} &
%   \ctx^3 &= \PHstate{f_1, a_1, c_1} \text{ and } P_2 = \PHobj{a_1}{a_0} \\
%   \ctx^2 &= \PHstate{f_1, a_1, c_0} \text{ and } P_3 = \PHobj{c_0}{c_1} &
%   \ctx^4 &= \PHstate{f_1, a_0, c_1} \text{ and } P_4 = \PHobj{c_1}{c_0}
% \end{align*}
% Given \pref{th:approxinf}, we have: $\forall i \in \segm{1}{4}, \muconcr_{\ctx^i}(P_i) \neq \emptyset$.
% Finally, after checking that it can be applied for $i = 3$ down to $i = 1$,
% \pref{thm:ordered-ua} gives: $\uconcr(\w) \neq \emptyset$.
% Thus the model is functional.

\begin{figure}[tp]
  \centering
  \begin{tikzpicture}[aS]
    % STEP 1
    \node[Aobj] (a01) {$\PHobj{a_0}{a_1}$};
    \node[Asol,below of=a01] (a01s) {};

    \node[AprocPrio,below of=a01s] (fc10) {$fc_{10}$};
    \node[AsolPrio,below of=fc10] (fc10s) {};

    \node[Aproc,below left of=fc10s] (f1) {$f_1$};
    \node[Aobj,below of=f1] (f11) {$\PHobj{f_1}{f_1}$};
    \node[Asol,below of=f11] (f11s) {};

    \node[Aproc,below right of=fc10s] (c0) {$c_0$};
    \node[Aobj,below of=c0] (c00) {$\PHobj{c_0}{c_0}$};
    \node[Asol,below of=c00] (c00s) {};

    \path
    (a01) edge (a01s)
    (a01s) edge (fc10)
    (fc10) edge[aSPrio] (fc10s)
    (fc10s) edge (f1) edge (c0)

    (f1) edge (f11)
    (f11) edge (f11s)

    (c0) edge (c00)
    (c00) edge (c00s)
    ;

    % STEP 2
    \node[Aobj,right of=a01,node distance=3.5cm] (a10) {$\PHobj{a_1}{a_0}$};
    \node[Asol,below of=a10] (a10s) {};

    \node[Aproc,below of=a10s] (c1) {$c_1$};
    \node[Aobj,below right of=c1] (c01) {$\PHobj{c_0}{c_1}$};
    \node[Asol,below of=c01] (c01s) {};
    \node[Aobj,below left of=c1] (c11) {$\PHobj{c_1}{c_1}$};
    \node[Asol,below of=c11] (c11s) {};

    \node[AprocPrio,below of=c01s] (fa11) {$fa_{11}$};
    \node[AsolPrio,below of=fa11] (fa11s) {};

    \node[Aproc,below left of=fa11s] (a1) {$a_1$};
    \node[Aobj,below of=a1] (a11) {$\PHobj{a_1}{a_1}$};
    \node[Asol,below of=a11] (a11s) {};

    \node[Aproc,below right of=fa11s] (f1) {$f_1$};
    \node[Aobj,below of=f1] (f11) {$\PHobj{f_1}{f_1}$};
    \node[Asol,below of=f11] (f11s) {};

    \path
    (a10) edge (a10s)
    (a10s) edge (c1)

    (c1) edge (c01) edge (c11)
    (c01) edge (c01s)
    (c11) edge (c11s)
    (c01s) edge (fa11)

    (fa11) edge[aSPrio] (fa11s)
    (fa11s) edge (f1) edge (a1)

    (f1) edge (f11)
    (f11) edge (f11s)

    (a1) edge (a11)
    (a11) edge (a11s)
    ;

    % STEP 3
    \node[Aobj,right of=a10,node distance=4cm] (ta01) {$\PHobj{a_0}{a_1}$};
    \node[Asol,below of=ta01] (ta01s) {};

    \node[AprocPrio,below of=ta01s] (tfc10) {$fc_{10}$};
    \node[AsolPrio,below of=tfc10] (tfc10s) {};

    \node[Aproc,below left of=tfc10s] (tf1) {$f_1$};
    \node[Aobj,below of=tf1] (tf11) {$\PHobj{f_1}{f_1}$};
    \node[Asol,below of=tf11] (tf11s) {};

    \node[Aproc,below right of=tfc10s] (tc0) {$c_0$};
    \node[Aobj,below of=tc0] (tc10) {$\PHobj{c_1}{c_0}$};
    \node[Asol,below of=tc10] (tc10s) {};
    \node[Aobj,right of=tc10] (tc00) {$\PHobj{c_0}{c_0}$};
    \node[Asol,below of=tc00] (tc00s) {};

    \node[Aproc,below of=tc10s] (ta0) {$a_0$};
    \node[Aobj,below of=ta0] (ta00) {$\PHobj{a_0}{a_0}$};
    \node[Asol,below of=ta00] (ta00s) {};

    \path
    (ta01) edge (ta01s)
    (ta01s) edge (tfc10)
    (tfc10) edge[aSPrio] (tfc10s)
    (tfc10s) edge (tf1) edge (tc0)

    (tf1) edge (tf11)
    (tf11) edge (tf11s)

    (tc0) edge (tc00)
    (tc00) edge (tc00s)
    (tc0) edge (tc10)
    (tc10) edge (tc10s)

    (tc10s) edge (ta0)
    (ta0) edge (ta00)
    (ta00) edge (ta00s)
    ;
  \end{tikzpicture}
  \caption{%
  \label{fig:metazoan-sa}%
    The three successive saturated graphs of local causality
    of the PH model in \pref{fig:metazoan-ph}
    for the objective sequence
    $\w = \PHobj{a_0}{a_1} \concat \PHobj{a_1}{a_0} \concat \PHobj{a_0}{a_1}$
    and the initial context
    $\ctx = \PHstate{a_0, c_0, f_1, fc_{10}, fa_{10}}$.
    The (left) graph allows to check objective $\PHobj{a_0}{a_1}$
    from the initial context $\ctx$.
    The (middle) graph is for the objective $\PHobj{a_1}{a_0}$
    and the context $\PHstate{a_1, c_0, f_1, fc_{10}, fa_{10}}$.
    The (right) graph is for the last objective, $\PHobj{a_0}{a_1}$,
    and the context $\PHstate{a_0, c_1, f_1, fc_{10}, fa_{11}}$.
%     Rectangular nodes containing a single process are elements in $\Proc$,
%     borderless nodes containing a couple of processes are elements in $\Obj$
%     and circle nodes are elements in $\Sol$.
%     \pref{th:approxinf} is inconclusive on this example as edge $(ab_{11}, \{ a_1, b_1 \}) \in \Proc \times \Sol$ (here represented with a double line) is not coherent (\pref{def:coherent}).
%     Indeed, $a_0 \in \Proc$ is a child of $\{ a_1, b_1 \}$, but $a_0 \neq a_1$ (and the same also stands for $b_0$).
  }
\end{figure}

\subsection{Large-scale Application: T-cell receptor}
\label{ssec:ex-tcrsig}

In order to support the scalability and applicability of our under-approximation of reachability, we
apply our new approach to the analysis of a large-scale model of the T-cell receptor (TCR)
signalling pathway \cite{tcrsig94}.
This model gathers 94 interacting components and is specified as a Boolean network.
The under-approximation presented in this paper has been implemented in the existing Pint
software\footnote{Pint is freely available at \url{http://loicpauleve.name/pint}.}.

### 2014-12-18 parts/intro.tex
add TODOs after review
refactor successive reachability in metazoan example

The main additions in this paper compared to \cite{FPMR13-CS2Bio} are
two theorems, the first
stating that any Process Hitting model with any number of classes of priorities can be represented
with only two classes of priorities,
and the other giving a method to cope with successive reachability,
both extending the application range of the static analysis we developed.

>>>>

The main addition in this paper compared to \cite{FPMR13-CS2Bio} is
%two theorems, the first
\pref{thm:ordered-ua} that gives a method to cope with successive reachability,
and obtain more conclusive results in the case of several successive objectives.
Two other additions are 
\pref{th:bisimPHP} which states that any Process Hitting model with any number of classes of priorities
can be represented with only two classes of priorities,
thus extending the scope of our method to any possible Process Hitting model,
and \todo{def extraction}
which allows to extract a concrete scenario when our static analysis method
is conclusive.
%Both aim at extending the application range of the static analysis we developed.
